---
title: Do Formula 1 Cars get faster over time?
author: Lukas Heck
date: '2019-11-01'
slug: f1-racing-over-time
categories: [R]
tags:
  - rstats
  - r
  - data
  - ggplot2
  - visuzalisation
---


#### How much faster do F1 tracks get over time?

With the help of the F1 Dataset provided at [Kaggle](https://www.kaggle.com/cjgdev/formula-1-race-data-19502017), we are going to explore how much faster F1 cars have gotten over time.

The datasets we are going to use are `races`, which includes data about the specific races, like name of the circuit, year of the GP. 
`laps` provides lap times for each GP, each driver and each round in that GP. The last dataset we are going to use, `results` includes the results for each GP and driver. 
The strategy is to find different variables of interest from the differnt dataframes and bind them using the handy `raceId`, that is provided in each dataset.

##### First step: Load and wrangle the data:



##### Setup and load packages:

```{r, message=FALSE, warning = FALSE}

library(tidyverse) # for... everything?
library(ggthemes)  # to make ggplot charts even more beautiful
library(lubridate) # to manipulate race-times
library(viridis)
library(plotly)


```

##### Include and inspect data:

```{r}
races <- read.csv("data/f1/races.csv")
head(races)
```



```{r, message=FALSE, warning = FALSE}
laps <- read.csv("data/f1/lapTimes.csv")

laps$raceId <- as.character(laps$raceId) 
laps$sec_round <- laps$milliseconds / 1000 # Create a variable for seconds, for easier understanding and interpretability.


head(laps)

 laps %>%
  filter(position < 6) %>%
  mutate(position = as.character(position)) %>%
  ggplot(aes(sec_round, fill=position, colour = position)) + geom_density(alpha = 0.1) + xlim(50,200) + theme_economist()

```


Inspecting the overall race times for positions 1 to 5, there is the smallest tendency to the left for the first two positions, but
overall, the variance is very large.





##### Create a Dataset with races and mean lap times.

```{r, message=FALSE, warning = FALSE}

avg_lap <- laps %>%
  group_by(raceId) %>%
  filter(lap == round(mean(laps$lap, na.rm = T))) %>%
  summarise(mean_race = mean(sec_round, na.rm = T))  # Calculate the mean lap time per race for a mean lap


races$raceId <- as.character(races$raceId)
f1 <- left_join(races, avg_lap,by = c("raceId")) %>% select("raceId", "name", "mean_race", "year" )

f1$name2 <- str_replace_all(f1$name, "Grand Prix", "")

head(f1)
```

Next step: add the `result` dataset with the fastest lap time.

```{r, message=FALSE, warning = FALSE}

res <- read.csv("data/f1/results.csv")
res$raceId <- as.character(res$raceId)
res$fastest <- lubridate::ms(res$fastestLapTime)
res$fastest <- as.duration(res$fastest)
res$fastest <- as.numeric(res$fastest)
res2 <- res %>%
  filter(!(is.na(fastestLapTime)) & fastest > 1) %>%
  group_by(raceId) %>%
  summarise(mean_fastest = mean(fastest),na.rm = T)
f1 <- left_join(f1,res2,by="raceId")

f1 <- f1 %>% filter(year > 2004)




```

Using `lubridate`, the `fastestLapTime` column is converted into a numeric value of seconds:
```{r}
res[1,] %>% select(fastest,fastestLapTime)
```


##### Exploration and Graphical Analysis



```{r, message=FALSE, warning = FALSE}


f1 %>% filter(name2=="Monaco ") %>%
ggplot( aes(mean_fastest )) + geom_density(alpha=.2, fill = "darkgreen") + geom_density(aes(x=mean_race), alpha=.2, fill= "red")  + theme_economist() + 
  scale_fill_brewer(palette = "Set3")  + 
  theme(legend.position="bottom") +
  labs(title = "Distribution of Laptimes for the Monaco GP 2005 - 2017", subtitle = 
         "Red: Average lap times for average rounds, Green: Fastest laps", y="Density", x= "Lap Time") 

# f1  %>%
# ggplot( aes(mean_fastest )) + geom_density(alpha=.2, fill = "darkgreen") + geom_density(aes(x=mean_race), alpha=.2, fill= "red")  + theme_economist() + 
#   scale_fill_brewer(palette = "Set3")  + 
#   theme(legend.position="bottom") +
#   labs(title = "Distribution of Laptimes for the Monaco GP 2005 - 2017", subtitle = 
#          "Red: Average lap times for average rounds, Green: Fastest laps", y="Density", x= "Lap Time") +  facet_wrap(~name2,ncol=9) 
```




```{r, message=FALSE,warning=FALSE}
x <- ggplot(f1, aes(x = abs(year),
               y=mean_race,
               color = mean_race)) + 
                                    geom_point()+  
                                    geom_line(aes(y=mean_fastest), color="red")+
                                    facet_wrap(~name2,ncol=9) +
                                    theme_economist() + 
                                    labs(title = "Laptimes in F1 over time", subtitle = "Grouped by Grand Prix, the line represents the fastest laps, the dots represent average laptimes.",
                                         y="Mean lap time (in Seconds)", x= "Year") +
                                    theme(legend.position="bottom",
                                          axis.text.x = element_text(size = 5, angle=45),
                                          axis.text.y = element_text(size = 6),
                                          strip.text.x = element_text(size=8)) +
  scale_color_gradientn(name="",colours=rev(viridis::viridis(20))) +   guides(color=FALSE)   + theme(axis.line = element_line(colour = NA), 
    axis.ticks = element_line(linetype = "blank")) 

ggplotly(x)
```


```{r, message=FALSE, warning = FALSE}


f1 %>%
  filter(year > 2004 ) %>%
  ggplot(aes(x=abs(year), y=mean_race, color = mean_race)) + geom_smooth(method = "loess") + geom_jitter()  + 
      scale_x_continuous(limits = c(2005,2017),    
                         breaks  =  seq(2005,2018,by = 1)) +
  theme_economist() +   scale_color_gradient(low="#1c8000",high="red") +   guides(color=FALSE) +
  geom_smooth(aes(y = mean_fastest), fill="red") +  labs(title = "Laptimes in F1 over time",
                                                                    subtitle = "Blue = Average Laptime, Red = Fastest Lap",
                                                                    y="Mean lap time", x= "Year") 
f1 %>%
  filter(year > 2004 ) %>%
  ggplot(aes(x=abs(year), y=mean_race, color = mean_race)) + 
  geom_smooth(method = "loess", fill="blue", color = "blue", size=.5, alpha=.3) +
  geom_jitter()  + 
      scale_x_continuous(limits = c(2005,2017),    
                         breaks  =  seq(2005,2018,by = 1)) +
  theme_economist() +
  scale_color_gradient(low="#1c8000",high="red") + 
  guides(color=FALSE) +
  geom_line(aes(y = mean_fastest), fill="red", color = "red") +
  labs(title = "Laptimes in F1 over time", 
       subtitle = "Blue = Average Laptime, Red = Fastest Lap",
       y="Mean lap time", x= "Year") +
  facet_wrap(~name2,ncol=9) + 
  theme(axis.line = element_line(colour = NA), 
  axis.ticks = element_line(linetype = "blank"),
  axis.text.x = element_text(size = 5, angle=45),
  axis.text.y = element_text(size = 6),
  strip.text.x = element_text(size=8)) 


  


```

